Data Management and Domain Adjusting Strategies for Implementing Parallel SPH Method in JPUE Simulation

### 2016-01-08 Paper for submission.tex
cleaned up, ready to submit

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
\author{\IEEEauthorblockN{Zhixuan Cao, Abani Patra}
\IEEEauthorblockA{Department of MAE \\%Mechanical and Aerospace Engineering
 The University at Buffalo\\
 Buffalo, New York 14260\\
Email: zhixuanc@buffalo.edu, abani@buffalo.edu}
\and
\IEEEauthorblockN{Matthew Jones, Jaroslaw Zola}
\IEEEauthorblockA{Center for Computational Research\\
Department of CSE \\%Computer Science and Engineering
The University at Buffalo\\
Buffalo, New York 14260\\
Email: jonesm@buffalo.edu, \\ jzola@buffalo.edu}
%\and
%\IEEEauthorblockN{Jaroslaw Zola}
%\IEEEauthorblockA{The University at Buffalo\\
%Department of Computer Science and Engineering\\
% The University at Buffalo\\
% Buffalo, New York 14260\\
%Email:jzola@buffalo.edu}
}
% conference papers do not typically use \thanks and this command
% is locked out in conference mode. If really needed, such as for
% the acknowledgment of grants, issue a \IEEEoverridecommandlockouts
% after \documentclass

@@

However, most implementation of parallel SPH method presented to date are limited to standard SPH and benchmark problems like dam break, or relatively
simple scenarios like breaking-waves, floodings etc. Work on more complicated problems, such as eruption of volcano plume (figure \ref{fig:Plume}) which is essentially a mult-phase turbulent flow,

@@

In both SPHysics\cite{dominguez2011optimization} and DualSPHysics\cite{crespo2015dualsphysics}, The physical quantities of each particle (position, velocity, density…) are stored in arrays, and the particles (and the arrays with particle data) are reordered following the order of the cells. This has two advantages: 1) access pattern is more regular and more efficient, 2) it is ease to identify the particles that belong to a cell by using a range since the first particle of each cell is known. But adding, deleting and especially accessing of particles are cumbersome. Ferrari\cite{ferrari2009new} adopted linked lists using pointers so that particles can be deleted or added during the simulation. Storage problems caused by fix-size arrays are thereby also eliminated. We define C++ classes which contains all data of particle and bucket. As for the management of data, we adopt a hash table to store pointers to particles and buckets, which gives us not only flexibility of deleting and adding element, but also quicker access compared with linked list. Instead of using the "nature manner" to number particles, we adopt SFC based index to give each particle and background bucket an unique identifier -- a strategy known to preserve some locality at minimal cost. The SFC based numbering strategy is further extended to include time step information so that particles added at the same position but different time will have different identifiers. 

@@

\includegraphics[scale=0.34]{Work_flow}

@@

SPH is a meshfree, Lagrangian method. The domain is discretized by particles and the position of each particle is updated at every time step. The physical laws (such as conservation laws of mass, momentum and energy) written in the form of partial differential equations need to be transformed into the Lagrangian particle formalism of SPH. Using a kernel function that provides the weighted estimate of the field variables in the neighborhood of a discrete point (particle), the integral equations are evaluated as sums over neighbor particles. Only particles located within support of kernel function will interact. Thus, physical properties (density, velocity, internal energy...) are updated based on its neighbors. So a neighbor searching needs to be carried out before updating of physical properties. We use buckets which contain all particles associated with a sub-domain 
and are kept fixed in time during the entire simulation, to reduce search cost (since search can now be restricted to only neighboring domains).
Domain decomposition will be based on an SFC going through centroids of all buckets. A basic work flow of our SPH code is shown in figure \ref{fig:Work_flow}.\\

@@

The SFC \cite{sagan2012space} maps n-dimentional space to a one dimentional sequence. The standard procedure for obtainning SFC is: 

@@

\item All keys are sorted to form a sequence which is SFC. The SFC represents a curve passing through all particles (or centroid of buckets).

@@

SFC-based indexing scheme can guarantee uniqueness of particle identifier only in simple scenarios when particles are added once while setting up initial condition. In some situations, new particles need to be added while simulation. For example, new particles need to be added at the bottom of the eject vent for JPUE simulation. To distinguish particles added at the "same place" (the small area, all points within which will be mapped to the same $k$.) at different time steps, we extend the SFC-based key to time-dependent SFC based key by including date of birth of particles into the key. The time-dependent SFC based key can be written as: $[k,t]$, where $t$ is the time step. The map $h_n$ will become:

@@

The most basic data structure of SPH are particle, for problem description, and bucket, for neighbors searches and domain decomposition. Both are defined as classes in C++. Information that is contained in particle class can be categorized into six categories: ID(the key), affiliate(rank of the process that the particle belongs to), primitive variables (variables show up as unknowns in governing equations, e.g. density, velocity), secondary variables (physical properties that can be computed from primitive variables, they are stored to avoid repeated computations eg. pressure, temperature), flags (indicators, such as indicator for ghost particle and real particles, indicator for particles of different phases) and neighbor information (it is a vector of particle keys in our application). Similarly,  Information that is contained in  a bucket class can also be categorized into different categories: ID, affiliate, domain information (maximum and minimum coordinates, boundary information), flags (indicators, such as indicator for guest and non-guest, indicator for active and inactive), neighbor information (keys of 27 neighbor buckets including its own key) and owned particles. Objects defined based on these two classes are then accessed through hash tables.

@@

As discussed at the beginning of this section, implementation of SPH in more realistic scenarios requires dynamic memory management and flexible data access. One of the fundamental data structures that satisfy such requirement is hash table. Another option is B-tree. We adopt hash table. An implementation of B-tree under a similar situation for mesh based methods can be found in other papers (eg.\cite{patra2003data}).\\

@@

 %We should also do some experiment to optimize our hash table size.

@@

\caption{Non-uniform distribution of particles in the $[0,1]^n \times \textbf{T}$ space due to adding of new particles at a small portion of the domain, pointer to these new particles will be stored in external link list.}

@@

One natural way to hash time-dependent SFC based key $[k,t]$ is to convert the two elements in the key into one number taking $k$ as the higher digit and $t$ as the lower digit of the large number. However, for JPUE simulation, even though ghost particles for wall boundary condition and pressure boundary condition also need to be added during simulation, places for adding of these two types of ghost particles are previsouly empty area. Only ghost particles for eruption boundary condition will be successively added at the same place: bottom of the vent. That is to say, particles are distributed non-uniformly in the $[0,1]^n \times \textbf{T}$ space as shown in figure \ref{fig:Particle_adding_with_link}. To avoid non-uniform, very sparse hash table and conserve locality of SFC, we only plug the first number, $k$, of the key, $[k,t]$, into the hashing function, equation (\ref{eq:hash_function}). 

@@

  \caption{Computational Work Load for Each Type of Particle}

@@

    Particles & NS & UPME & UPD & UPP &VF &WBC &Total\\

@@

Figure \ref{fig:SFC_domain_decomposition} shows how domain is decomposed based on partition of SFC of buckets. The particles are automatically split into several groups along with buckets that contain them. As SFC of buckets is a curve in the three dimensional space, partition of this curve will automatically lead to 3D domain decomposition. A 2D domain decomposition based on SFC of footprint buckets projected by three dimensional buckets was adopted by Kumar et al \cite{kumar2013parallel}. A comparison of the scalability of these two schemes (see figure \ref{fig:2D_vs_3D_efficiency}) confirms that 3D domain decomposition is a better choice when processes number is larger. 

@@

As some of the neighbor particles reside in other partitions. A set of  particles and buckets are used to synchronize data across partitions. To minimize communications, data is synchronized only where needed, using non-blocking MPI communications. 

@@

\includegraphics[scale=0.36]{Work_flow_adjust}

@@

We propose a simple strategy to add such feature in our code with low computational cost. We add a scan function to monitor the most outside layer of the domain. When the ejected fluid reaches the boundary of the current domain, ghost particles (for pressure boundary condition) will be turn to real particles and then add new ghost particles for pressure boundary condition. The original work flow (see figure \ref{fig:Work_flow}) is modified to enable such a feature (see figure \ref{fig:Work_flow_adjust}).\\
This is implemented by using a involved-flag, in the particle class. Particles are categorized into three groups based on the value of the involved-flag: involved (involved-flag=2), potentially involved (involved-flag=1), and not involved (involved-flag=0). The involved particles are particles that have already been affected by the mixing. The potentially involved particles are particles that have not been involved in mixing but are adjacent to involved particles and will thus be be involved in the near future. %These not involved particles are particles that far away from the eject source and are still at the initial state. For these not involved particles, it is meaningless to update its physical properties. As a consequence, searching for neighbors and data communitation for these particles are also unnecessary.

@@

As simulation progresses, the ejected fluid will reach larger area and more and more particles will be influenced. When originally stationary air is influenced by erupted material, the mass fraction of the erupted material will increase from zero to a positive value. So we can determine whether a particle is involved or not based on whether the mass fraction of that particle is larger than a given threshold ($10^{-5} $ in our simulation). Other physical properties, like velocity, can also serve as alternative "swich criteria".\\

@@

Our target here is to develop data management and parallelization strategies for more complicated implementations of SPH which demand quick and flexible data access, deletion and addition. Thus, the test problem used should have such requirements. Simulations of a volcanic eruption, which is essentially a muti-phase, turbulent, ejection mixing flow accompanied with microphysics processes will place these demands on the framework. Thus, we adopt a two phase volcanic plume model\cite{suzuki2005numerical} as our test problem. In this model, one phase is air while another phase is ejected material.

@@

Experiments have been carried out on the computational cluster of Center for Computational Research (CCR) at Buffalo. Two types of processors -- 8 core L5630 running at 2.13GHz clock rate with 3GB memory per core on a Mellanox network and 12 core E5645 running at 2.40GHz clock rate with 4GB memory per core on a Q-Logic network. Each node comprised of two sockets with two of these processors. Memory and level 3 cache are shared on each node.

@@

\subfloat[A plume from Sakurajima Volcano eruption\cite{PlumePhoto}]{\includegraphics[width=3.8cm,height=4.5cm]{plume_photo}}

@@

\subfloat[Contour of mass fraction of erupted material at 290 seconds]{\includegraphics[width=3.8cm,height=4.5cm]{Plume_simulation}}

@@

Hashtables with external link list are adopted for accessing particles and buckets data. Based on weighted particle work load, a dynamic load balance strategy is developed by checking load balance and redecomposing the domain at an optimize interval. The performance of the code was further improved to several times faster by adjusting computational domain according to progress of simulation. 
Scalability tests on our code shew acceptable strong scalability and good weak scalability. 3D domain decomposition provided better strong scalability than 2D domain decomposition when number of processes is larger.

>>>>

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
\author{\IEEEauthorblockN{Zhixuan Cao, Abani Patra}
\IEEEauthorblockA{Department of MAE \\%Mechanical and Aerospace Engineering
 The University at Buffalo\\
 Buffalo, New York 14260\\
Email: zhixuanc@buffalo.edu, \\ abani@buffalo.edu}
\and
\IEEEauthorblockN{Matthew Jones, Jaroslaw Zola}
\IEEEauthorblockA{Center for Computational Research, \\
Department of CSE \\%Computer Science and Engineering
The University at Buffalo\\
Buffalo, New York 14260\\
Email: jonesm@buffalo.edu, \\ jzola@buffalo.edu}
%\and
%\IEEEauthorblockN{Jaroslaw Zola}
%\IEEEauthorblockA{The University at Buffalo\\
%Department of Computer Science and Engineering\\
% The University at Buffalo\\
% Buffalo, New York 14260\\
%Email:jzola@buffalo.edu}
}
% conference papers do not typically use \thanks and this command
% is locked out in conference mode. If really needed, such as for
% the acknowledgment of grants, issue a \IEEEoverridecommandlockouts
% after \documentclass

@@

However, most implementations of parallel SPH method presented to date are limited to standard SPH and benchmark problems like dam break, or relatively
simple scenarios like breaking-waves, floodings etc. Work on more complicated problems, such as eruption of volcano plume (figure \ref{fig:Plume}) which is essentially a multiphase turbulent flow,

@@

In both SPHysics\cite{dominguez2011optimization} and DualSPHysics\cite{crespo2015dualsphysics}, The physical quantities of each particle (position, velocity, density…) are stored in arrays, and the particles (and the arrays with particle data) are reordered following the order of the cells. This has two advantages: 1) access pattern is more regular and more efficient, 2) it is ease to identify the particles that belong to a cell by using a range since the first particle of each cell is known. But adding, deleting and especially accessing of particles are cumbersome. Ferrari\cite{ferrari2009new} adopted linked lists using pointers so that particles can be deleted or added during the simulation. Storage problems caused by fix-size arrays are thereby also eliminated. We define C++ classes which contains all data of particle and bucket. As for the management of data, we adopt hash tables to store pointers to particles and buckets, which give us not only flexibility of deleting and adding element, but also quicker access compared with linked list. Instead of using the "nature manner" to number particles, we adopt SFC based index to give each particle and background bucket an unique identifier -- a strategy known to preserve some locality at minimal cost. The SFC based numbering strategy is further extended to include time step information so that particles added at the same position but different time will have different identifiers. 

@@

\includegraphics[scale=0.33]{Work_flow}

@@

SPH is a meshfree, Lagrangian method. The domain is discretized by particles and the position of each particle is updated at every time step. The physical laws (such as conservation laws of mass, momentum and energy) written in the form of partial differential equations need to be transformed into the Lagrangian particle formalism of SPH. Using a kernel function that provides the weighted estimation of the field variables in the neighborhood of a discrete point (particle), the integral equations are evaluated as sums over neighbor particles. Only particles located within support of kernel function will interact. Thus, physical properties (density, velocity, internal energy...) associated to the particle are updated based on its neighbors. So a neighbor searching needs to be carried out before updating of physical properties. We use buckets which contain all particles associated with a sub-domain 
and are kept fixed in time during the entire simulation, to reduce searching cost (since search can now be restricted to only neighboring domains).
Domain decomposition will be based on a SFC going through centroids of all buckets. A basic work flow of our SPH code is shown in figure \ref{fig:Work_flow}.\\

@@

The SFC \cite{sagan2012space} maps a n-dimentional domain to a one dimentional sequence. The standard procedure for obtainning SFC is: 

@@

\item All keys are sorted to form a sequence which is SFC. The SFC represents a curve passing through all particles (or centroids of buckets).

@@

SFC-based indexing scheme can guarantee uniqueness of particle identifier only in simple scenarios when particles are added once while setting up initial condition. In some situations, new particles need to be added while simulation. For example, new particles need to be added at the bottom of the eject vent for JPUE simulation (see figure \ref{fig:Particle_adding_with_link}). To distinguish particles added at the "same place" (the small area, all points within which will be mapped to the same $k$.) at different time steps, we extend the SFC-based key to time-dependent SFC based key by including date of birth of particles into the key. Then the time-dependent SFC based key can be written as: $[k,t]$, where $t$ is the time step. The map $h_n$ will become:

@@

The most basic data structure of SPH are particle, for problem description, and bucket, for neighbors searches and domain decomposition. Both are defined as classes in C++. Information that is contained in particle class can be categorized into six categories: ID(the key), affiliate(rank of the process that the particle belongs to), primitive variables (variables show up as unknowns in governing equations, e.g. density, velocity), secondary variables (physical properties that can be computed from primitive variables, they are stored to avoid repeated computations eg. pressure, temperature), flags (indicators, such as indicator for ghost particle and real particles, indicator for particles of different phases) and neighbor information (it is a vector of particle keys in our application). Similarly,  Information that is contained in  a bucket class can also be categorized into different categories: ID, affiliate, domain information (maximum and minimum coordinates, boundary information), flags (indicators, such as indicator for guest and non-guest, indicator for active and inactive), neighbor information (keys of 27 neighbor buckets including its own key) and owned particles (it is a vector of particle keys in our application). Objects defined based on these two classes are then accessed through hash tables.

@@

As discussed at the beginning of this section, implementation of SPH in more realistic scenarios requires dynamic memory management and flexible data access. One of the fundamental data structures that satisfy such requirement is hash table. Another option is B-tree. We adopt hash table in our code. An implementation of B-tree under a similar situation for mesh based methods can be found in other papers (eg.\cite{patra2003data}).\\

@@

 %We should also do some experiment to optimize our hash table size. In addition, Dr. Patra's results might no longer reliable due to quick development in CPU architechture, such as changes in the size of caches.

@@

\caption{Non-uniform distribution of particles in the $[0,1]^n \times \textbf{T}$ space due to adding of new particles only at a small portion of the whole domain. These particles are added at different time steps. Pointers to these new particles will be stored in external link lists.}

@@

One natural way to hash time-dependent SFC based key $[k,t]$ is to convert the two elements in the key into one number taking $k$ as the higher digit and $t$ as the lower digit of the large number. However, for JPUE simulation, even though ghost particles for wall boundary condition and pressure boundary condition also need to be added during simulation, places for adding of these two types of ghost particles are previsouly empty area (see figure \ref{fig:Domain_adjusting}). Only ghost particles for eruption boundary condition will be successively added at the same place: bottom of the vent. That is to say, particles are distributed non-uniformly in the $[0,1]^n \times \textbf{T}$ space as shown in figure \ref{fig:Particle_adding_with_link}. To avoid non-uniform, very sparse hash table and conserve locality of SFC, we only plug the first number, $k$, of the key, $[k,t]$, into the hashing function, equation (\ref{eq:hash_function}). 

@@

  \caption{Computational Work Load for Different Types of Particles}

@@

    Types & NS & UPME & UPD & UPP &VF &WBC &Total\\

@@

Figure \ref{fig:SFC_particles_buckets_partition} shows how domain is decomposed based on partition of SFC of buckets. The particles are automatically split into several groups along with buckets that contain them. As SFC of buckets is a curve in the three dimensional space, partition of this curve will automatically lead to 3D domain decomposition. A 2D domain decomposition based on SFC of footprint buckets projected by three dimensional buckets was adopted by Kumar et al \cite{kumar2013parallel}. A comparison of the scalability of these two schemes (see figure \ref{fig:2D_vs_3D_efficiency}) confirms that 3D domain decomposition is a better choice when processes number is larger. 

@@

As some of the neighbor particles reside in other partitions. A set of "guest" particles and buckets are used to synchronize data across partitions. To minimize communications, data is synchronized only where needed, using non-blocking MPI communications. 

@@

\includegraphics[scale=0.32]{Work_flow_adjust}

@@

We propose a simple strategy to add such feature in our code with low computational cost. We add a scan function to monitor the outermost layer of the domain. When the ejected fluid reaches the boundary of the current domain, ghost particles (for pressure boundary condition) will be turn to real particles and then add new ghost particles for pressure boundary condition. The original work flow (see figure \ref{fig:Work_flow}) is modified to enable such a feature (see figure \ref{fig:Work_flow_adjust}).\\
This is implemented by using a involved-flag, in the particle class. Particles are categorized into three groups based on the value of the involved-flag: involved (involved-flag=2), potentially involved (involved-flag=1), and not involved (involved-flag=0). The involved particles are particles that have already been affected by the mixing. The potentially involved particles are particles that have not been involved in mixing but are adjacent to involved particles and will thus be involved in the near future. %These not involved particles are particles that far away from the eject source and are still at the initial state. For these not involved particles, it is meaningless to update its physical properties. As a consequence, searching for neighbors and data communitation for these particles are also unnecessary.

@@

As simulation progresses, the ejected fluid will reach larger area and more and more particles will be influenced. When originally stationary air is influenced by erupted material, the mass fraction of the erupted material will increase from zero to a positive value. So we can determine whether a particle is involved or not based on whether the mass fraction of that particle is larger than a given threshold ($10^{-5} $ in our simulation). Other physical properties, like velocity, can also serve as alternative "swich criterias".\\

@@

Our target here is to develop data management and parallelization strategies for more complicated implementations of SPH which demand quick and flexible data access, deletion and addition. Thus, the test problem used should have such requirements. Simulations of a volcanic eruption, which is essentially a multi-phase, turbulent, ejection mixing flow accompanied with microphysics processes will place these demands on the framework. Thus, we adopt a two phase volcanic plume model\cite{suzuki2005numerical} as our test problem. In this model, one phase is air while another phase is ejected material.

@@

Experiments have been carried out on the computational cluster of Center for Computational Research (CCR) at Buffalo. Two types of processors -- 8 core L5630 running at 2.13GHz clock rate with 3GB memory per core on a Mellanox network and 12 core E5645 running at 2.40GHz clock rate with 4GB memory per core on a Q-Logic network. Each node comprised of two sockets with several of these processors. Memory and level 3 cache are shared on each node.

@@

\subfloat[A plume from Sakurajima Volcano eruption\cite{PlumePhoto}]{\includegraphics[width=3.8cm,height=4.4cm]{plume_photo}}

@@

\subfloat[Contour of mass fraction of erupted material at 290 seconds]{\includegraphics[width=3.8cm,height=4.4cm]{Plume_simulation}}

@@

Hashtables with external link list are adopted for accessing particles and buckets data. Based on weighted particle work load, a dynamic load balance strategy is developed by checking load balance and redecomposing the domain at an optimized interval. The performance of the code was further improved to several times faster by adjusting computational domain according to progress of simulation. 
Scalability tests on our code shew acceptable strong scalability and good weak scalability. It was also demonstrated that 3D domain decomposition provided better strong scalability than 2D domain decomposition when number of processes is larger.
